Parameter Inference\\with Bifurcation Diagrams

### 2021-05-22 docs/article.tex

\title{Inference of Bifurcations\\with Differentiable Continuation}
\author{
        Gregory Szep\\King's College London\\London, WC2R 2LS\\
        \texttt{gregory.szep@kcl.ac.uk}
    \And
        Attila Csik\'asz-Nagy\\King's College London\\London, WC2R 2LS\\
        \texttt{attila.csikasz-nagy@kcl.ac.uk}
    \And
        Neil Dalchau\\Microsoft Research Cambridge\\Cambridge, CB1 2FB\\
        \texttt{ndalchau@microsoft.com}
}

\begin{document}
\input{notation}
\maketitle

\begin{abstract}
    In this work we propose a gradient-based semi-supervised approach for matching target bifurcations with parameterised differential equation models. The cost function contains a supervised error term that is minimal when predicted bifurcations match the targets and an unsupervised bifurcation measure which acts as a proxy for the probability of a bifurcation, yielding non-zero gradients in non-bifurcating parameter regimes. The calculation of gradients with respect to parameters shares the same computational complexity as deflated pseudo-arclength continuation used to calculate the bifurcation diagram. Our implementation leverages automatic differentiation methods in Julia and we demonstrate model synthesis with minimal models which explore the space of saddle-node and pitchfork diagrams and the genetic toggle switch from synthetic biology. Furthermore, the cost landscape allows us to  organise models in terms of topological and geometric equivalence.
\end{abstract}

@@

Backpropagation through differential equation solves has been a breakthrough over the past couple of years \cite{Chen2018NeuralEquations,Rackauckas2019DiffEqFlux.jl-AEquations} that enabled scalable parameter inference for differential equations. Determining what model parameters should be for a given set of observations, in the setting of biology and engineering, are known as inverse problems \cite{Abdulla2009InverseBiology}.

Optimisation targets, however, have mostly been expressed in the spatio-temporal domain. Acquisition of such data can be costly and can often contain over-constraining information generated by processing steps or the measurement device rather than the state variables that underpin the observed mechanism. In microscopy, for example, data is often reported in arbitrary fluoresce units allowing the observer to shift and scale data arbitrarily. Furthermore, such data may also not contain sufficient information about dynamical transients in order to identify kinetic parameters. The emerging picture suggests that identification of the qualitative behaviour -- the bifurcation diagram -- should precede any attempt at inferring kinetic parameters \cite{Stumpf2019ParameterBifurcations}. Techniques for back-propagating through implicit equation solvers have also been developed \cite{Look2020DifferentiableLayers,Bai2019DeepModels} although to the best of the authors' knowledge have not been applied to bifurcation diagrams at the time of writing this paper.

In this work we propose a gradient-based semi-supervised approach that focuses on fitting high-level qualitative constraints, defined by state space structures, rather than kinetics. Drawing inspiration from implicit solvers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients we find that their computation shares the same  complexity as the algorithm used to calculate the bifurcation diagram. We use a predictor-corrector method called deflated pseudo-arclength continuation \cite{Farrell2016TheDiagrams,Veltz2019PseudoArcLengthContinuation.jl}, originally developed for partial differential equations. In the case of partial differential equations the computational complexity of calculating a single bifurcation diagram is not bounded since superpositions of localised solutions may give rise to uncountably many branches \cite{Avitabile2010ToEquation}. The complexity of computing a single branch, however, is bounded by the complexities of the chosen eigenvalue solver and corrector, and further decreased by adaptive stepping procedures \cite{Aruliah2016AlgorithmContinuation}.

We find that the cost function landscape contains basins that not only allow us to synthesise models with a desired bifurcation structure but also allow us to organise models in terms of topological and geometric equivalence. We discuss the relevance of this in model selection. In summary, our paper has the following main contributions:

\begin{itemize}
    \item Defined a differentiable semi-supervised cost function for encouraging and placing co-dimension one bifurcations to user-specified target locations
    \item Implementation of method in Julia package \texttt{FluxContinuation.jl} leveraging automatic differentiation tooling in \texttt{Flux.jl}
    \item Leveraging the cost landscape for a novel way of organising differential equation models in terms of geometric and topological equivalence
\end{itemize}

\subsection{Related Works | \textit{in progress}}

\paragraph{Bifurcation control} and \textit{inverse} bifurcation analysis literature applies control theory to enable design and identification of dynamical systems with qualitative constraints. The first works involved the calculation of the minimal distances to bifurcation manifolds \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations}. Those optimisations did not involve closed form expressions and the challenge of finding the manifolds to being with was left to random sampling.

\paragraph{Matching vector fields} From a dynamical systems perspective a given bifurcation curve constrains a state-vector field geometry. A set of target bifurcations may also be represented as a vector field flow and thus inferring bifurcations becomes a problem in matching flows in vector sub-spaces. Note that this would involve matching direction but not the magnitude of the flow.

\begin{itemize}
    \item smooth and match estimator methods \cite{Ranciati2017BayesianParameters}
    \item robots learning limit cycles \cite{Khadivar2021LearningBifurcations}
\end{itemize}

\paragraph{Non-differentiable methods} 
A large body of work is dedicated to finding conditions for fold-bifurcations, multistability and other

\begin{itemize}
    \item bifurcation inference using Mixed Integer Nonlinear Programming \cite{Otero-Muras2018Optimization-basedModels}
    \item Random sampling and genetic algorithms \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}
\end{itemize}

\subsection{Preliminaries}
Suppose we parameterise a set of differential equations for states $u\in\Reals^N$ with a function $\rates$ in an unknown parameter space $\theta\in\Reals^M$. We would like these differential equations to obey a set of target bifurcations $\targets:=\{p_1\dots p_K\}$ along a known bifurcation parameter $p\in\Reals$. Let the differential equations be defined as
\begin{align}
	\partial_t u=\rates(z)
	\qquad\mathrm{where}\quad z:=(u,p)\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}
For a given set of parameters $\theta$ one could compute the set of predicted bifurcations $\predictions(\theta)$ using parameter continuation methods \cite{Veltz2019PseudoArcLengthContinuation.jl,Farrell2016TheDiagrams}. Our goal is to find optimal parameters $\theta^*$ that match predictions $\predictions(\theta^*)$ to specified targets $\targets$. We must design a suitable cost function $\loss$ so that
\begin{align}
    \theta^* := \mathrm{argmin}_{\theta} \loss(\theta|\targets)
    \label{eq:optimal-theta}
\end{align}
The optimal $\theta^*$ is not expected to always be unique, but is in general a manifold representing the space of qualitatively equivalent models. The cost function $\loss$ must have some distance measure between the two sets $\predictions(\theta)$ and $\targets$ with a bias that encourages equivalent cardinality $|\predictions(\theta)|=|\targets|$. This is especially important in the case where there are no predictions $|\predictions(\theta)|=0$.

For clarity, we guide the reader through the methods with the following minimal models that explore the space of saddle-nodes $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchforks $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$. These minimal models are shown with targets $\targets$ in Figure \ref{fig:saddle-node} and Figure \ref{fig:pitchfork} respectively. These figures show that the determinant $\Det=0$ at the bifurcations and that its directional derivative along the bifurcation curve $\frac{d}{ds}\Det$ is finite and non-zero, at least in the non-degenerate case. This gives us our first hint of what should be optimised to increase the number of predictions $|\predictions(\theta)|$.

\begin{figure}
\centering
\includegraphics[width=6cm]{saddle-node}
\includegraphics[width=6cm]{pitchfork}
\caption{Saddle-node model $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchfork model $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$ with $\theta=(5/2,-1),(1/2,-1)$ and targets $\targets=\{-1,1\},\{0\}$ respectively. Determinant in red. Lighter shades indicate the determinant crossing zero for unstable solutions}
\label{fig:minimal-models}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=7cm]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(z)$ and determinant $\Det$ for two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars}
    \label{fig:bifurcation-measure}
\end{figure}

In order for predicted bifurcations $p(\theta)\in\predictions$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. We expect this term to be some mean over the targets $\targets$ and predictions $\predictions(\theta)$. We choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target in cases where $|\predictions|\geq|\targets|$. This term only zero when each target is matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:supervised-term}
\end{align}
We can see from Figures \ref{fig:saddle-node} and \ref{fig:pitchfork} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$ in the non-degenerate case. Using these quantities we can define a positive semi-definite measure $\measure(z)$ of zero crossings in the determinant along the bifurcation curve; as shown in Figure \ref{fig:bifurcation-measure} this can be used as a proxy for the probability of a bifurcation at location $p(\theta)$. The total bifurcation measure $\Psi(\theta)$ is written
\begin{align}
    \Psi(\theta):=\frac{
        \int_{\rates(z)=0}\!
        \measure(z)\,\mathrm{d}z
    }{
        \int_{\rates(z)=0}\!\
        \mathrm{d}z
    }
    \quad\mathrm{where}\quad
    \measure(z):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{align}
Here we denote $\int_{\rates(z)=0}\mathrm{d}z$ as the sum of the line integrals in $z\in\Reals^{N+1}$ defined by the steady states $\rates(z)=0$. The measure $\measure(z)=1$ at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in $\frac{d}{ds}\Det\rightarrow0$ and hence $\measure(z)\rightarrow0$ for each turning point. As $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(z)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(z)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the observed parameter region $p$ is densely packed with bifurcations. The $\measure(z)$ is also invariant under certain classes of transformations on $\rates(z)$ which do not change bifurcation locations, proofs of which are provided in Appendix \ref{appendix:measure}. The measure is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:cost}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that in principle while individual bifurcations $p(\theta)$ depend smoothly on $\theta$ the total number of predictions $|\predictions|$ does not. The gradient $\partial_{\theta}|\predictions|=0$ everywhere and is an infinite impulse at the exact locations where the number of predictions changes.

In practice optimisers are unlikely to hit those values of $\theta$ and we can safely drop the dependency. We may now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ and total measure $\Psi(\theta)$
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation curve $\rates(z)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:space-curve}.

The gradient of the bifurcation points $\frac{\partial p}{\partial\theta}$ is found by application of the implicit function theorem to a vector function $G_{\theta}:\Reals^{N+1}\rightarrow\Reals^{N+1}$ whos components represent the two constraints $\rates(z)=0$ and $\Det=0$. By following a similar strategy to that used by implicit layers \cite{Look2020DifferentiableLayers} we yield an $(N+1)\times M$ Jacobian representing a deformation field \cite{Jos2011OnSurface} for each $\theta$ direction. The gradient we are looking for becomes
\begin{align}
    \frac{\partial p}{\partial\theta} = -\hat{p}\cdot\left.\frac{\partial G_{\theta}}{\partial z}^{-1}
    \frac{\partial  G_{\theta}}{\partial\theta}\right|_{G_{\theta}(z)=0}
    \quad\mathrm{where}\quad
    G_{\theta}(z):=\begin{bmatrix}\rates(z)\\\Det\end{bmatrix}
\end{align}
Here $\hat{p}$ is a unit vector in $z\in\Reals^{N+1}$ that picks out the deformations along the $p$-direction. We can see this calculation involves inverting an $(N+1)\times(N+1)$ Jacobian $\frac{\partial G_{\theta}}{\partial z}$. The determinant of this Jacobian goes to zero when $\frac{d}{ds}\Det=0$. This further justifies our choice for measure $\measure(z)$ which discourages such cases.

The cost function is piece-wise smooth and differentiable with undefined gradients only in parameter regions where the number of predictions $|\predictions|$ changes. Given a set of solutions to $\rates(z)=0$ and locations $p(\theta)$ which additionally satisfy $\Det=0$, the gradient $\frac{\partial\loss}{\partial\theta}$ can evaluated using automatic differentiation methods.

\subsection{Implementation}
\label{section:implementation}
%/ move to appendix?
Here provide a high-level view of Algorithm \ref{alg:optimisation-loop}. We use \texttt{BifurcationKit.jl} \cite{Veltz2019PseudoArcLengthContinuation.jl} to calculate bifurcation diagrams, which requires the rate function $\rates$, its Jacobian $J_\theta$ which can be calculated using \texttt{ForwardDiff.jl} \cite{Revels2016Forward-ModeJulia}, and an initial guess $u_0$ at an initial parameter $p_0$. Let this algorithm be called by function \texttt{getSteadyStates} and return the steady states that form the integration region $\steadystates$ and a new guess $u_0$ that is the true steady state at $p_0$.

The algorithm also takes hyperparameters $\beta$ which will contain arclength step sizes, eigenvalue and Newton solver options and unsupervised correction factor $\lambda$. The adaptive hyperparameter update is done by \texttt{updateHyperparameters} and should depend on the current solutions $\steadystates$ and the targets $\targets$. The step sizes and solver options should adapt to the expected space of the solutions $\steadystates$ so that the waiting time between iterations is minimised. The unsupervised correction factor $\lambda$ should be large when there are no predicted bifurcations and decrease as the number of predicted bifurcations approaches the number of targets.

The evaluation of the cost gradient is done by \texttt{costGradient} using \texttt{ForwardDiff.jl} \cite{Revels2016Forward-ModeJulia} which has a large repository of chain rules that it uses to correctly evaluate derivatives. At the time of writing this paper the repository does not include the Leibniz rule \cite{Flanders1973DifferentiationSign} for implicitly defined line integrals, and we therefore define additional rules that implement the results shown in Appendix \ref{appendix:space-curve} and \ref{appendix:deformation}. The cost gradient is used by \texttt{updateParameters}, which can be any gradient-based optimiser such as ADAM or Momentum gradient decent, to update the parameters $\theta$.

In practice the bifurcation curve is only evaluated in a finite region $p\in\Omega$ which includes the targets $\targets$. We are now ready to write down the semi-supervised cost function for a bifurcation curve defined by $\rates(z)=0$ in an observed region $p\in\Omega$.

>>>>

\title{Parameter Inference\\with Bifurcation Diagrams}
\author{
        Gregory Szep\\King's College London\\London, WC2R 2LS\\
        \texttt{gregory.szep@kcl.ac.uk}
    \And
        Attila Csik\'asz-Nagy\\King's College London\\London, WC2R 2LS\\
        \texttt{attila.csikasz-nagy@kcl.ac.uk}
    \And
        Neil Dalchau\\Microsoft Research Cambridge\\Cambridge, CB1 2FB\\
        \texttt{ndalchau@microsoft.com}
}

\begin{document}
\input{notation}
\maketitle
\begin{abstract}
    Estimation of parameters in differential equation models can be achieved by applying learning algorithms to time-series data. However, sometimes it is only possible to measure qualitative changes of a system in response to a controlled condition. In dynamical systems theory, such change points are known as \textit{bifurcations} and lie on a function of the controlled condition called the \textit{bifurcation diagram}. In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. The cost function contains a supervised error term that is minimal when the model bifurcations match the specified targets and an unsupervised bifurcation measure which has gradients that push optimisers towards bifurcating parameter regimes. The gradients can be computed without the need to differentiate through the operations of the solver that was used to compute the diagram. We demonstrate parameter inference with minimal models which explore the space of saddle-node and pitchfork diagrams and the genetic toggle switch from synthetic biology. Furthermore, the cost landscape allows us to organise models in terms of topological and geometric equivalence.
\end{abstract}

@@

Inverse problems \cite{Abdulla2009InverseBiology} arise in biology and engineering in settings when the model is not fully known and the desire is to match model behaviour to a given set of observations. This helps systematically guide both model and experimental design. The collected observations, however, are often indirectly proportional to the variables defined in the model; in microscopy, for example, data are reported in arbitrary fluorescence units allowing the observer to shift and scale data arbitrarily. The model, on the other hand, describes the dynamics of concentrations of relevant proteins. Furthermore, the observed qualitative changes in response to changes in experimental conditions are more robust and reproducible across studies than the quantitative details. For example, several studies are likely to agree that the human immune system activates above a threshold concentration of a pathogen and deactivates at a lower threshold concentration, but may disagree on the exact quantities of the thresholds or the magnitudes of the immune response. Bifurcation theory provides us a framework for studying these qualitative changes in a manner that is independent of quantitative details. The emerging picture suggests that identification of the qualitative behaviour -- the bifurcation diagram -- should precede any attempt at inferring other properties of a system \cite{Stumpf2019ParameterBifurcations}.

Inferring the parameters of a model directly from a bifurcation diagram is difficult because it is not obvious how one should change the parameters to create a bifurcation. It could even be impossible for the model to bifurcate in the manner desired. Several approaches exist to place bifurcations to desired locations once a manifold is present \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations} yet typically resort to sampling techniques to search for them in the first place \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}. Progress has been made is cases where model structure and stability conditions are used to refine the search space \cite{Otero-Muras2018Optimization-basedModels,Otero-Muras2014ACurves} yet the resulting objectives are still not explicit in the bifurcation targets and also not differentiable. A scalable method for navigating the space of bifurcation diagrams would enable design of differential equations with high-level qualitative constraints. Furthermore one could begin organising models according to qualitatively distinct behaviours.

Back-propagation through differential equation solvers has been a breakthrough over the past couple of years \cite{Chen2018NeuralEquations,Rackauckas2019DiffEqFlux.jl-AEquations} that enabled scalable parameter inference for differential equations from trajectory data. Although one could use trajectory data to create the aforementioned qualitative constraints \cite{Ranciati2017BayesianParameters,Khadivar2021LearningBifurcations} this would entail over-constraining information originating from the kinetics and dynamical transients of the model. Furthermore, such data usually does not contain sufficient information about dynamical transients in order to identify kinetic parameters. Techniques for back-propagating through implicit equation solvers have also been developed \cite{Look2020DifferentiableLayers,Bai2019DeepModels} although to the best of the authors' knowledge have not been applied to bifurcation diagrams at the time of writing this paper.

In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. The bifurcation diagram encodes high-level qualitative information defined by state space structures, rather than kinetics. Drawing inspiration from implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients we find that their computation shares the same  complexity as the algorithm used to calculate the bifurcation diagram. The compute the diagram we use a predictor-corrector method called deflated pseudo-arclength continuation \cite{Farrell2016TheDiagrams,Veltz2019PseudoArcLengthContinuation.jl} developed for partial differential equations. In the case of partial differential equations the computational complexity of calculating a single bifurcation diagram is not bounded since superpositions of localised solutions may give rise to uncountably many branches \cite{Avitabile2010ToEquation}. The complexity of computing a single branch, however, is bounded by the complexities of the chosen eigenvalue solver and corrector, and further decreased by adaptive stepping procedures \cite{Aruliah2016AlgorithmContinuation}.

We find that the cost function landscape contains basins that not only allow us to synthesise models with a desired bifurcation diagram but also allow us to organise models in terms of topological and geometric equivalence. We discuss the relevance of this in model selection. In summary, our paper has the following main contributions:

\begin{itemize}
    \item An end-to-end differentiable method for searching for bifurcations and then matching them to user-specified target locations
    \item Implementation of the method in Julia package leveraging automatic differentiation tooling
    \item Leveraging the cost landscape for a novel way of organising differential equation models in terms of geometric and topological equivalence
\end{itemize}

\subsection{Preliminaries}

Suppose we collected observations along a scalar control condition $p\in\Reals$ and conclude that there are specific values of $p$ for which there are qualitative changes in system behaviour. Let $\targets$ be the set of those values and let us hypothesise that these transitions occur due to bifurcations in the dynamics that drive the underlying mechanism. Let us model the mechanism with a parameterised set of differential equations for states $u\in\Reals^N$ with a vector function $\rates$ in a parameter space $\theta\in\Reals^M$.

For the purposes of introducing this work, we will consider the simplest class of bifurcations known as \textit{co-dimension one} bifurcations, that do not include limit cycles. Therefore $\targets$ should contain conditions for which we hypothesise changes in multi-stable behaviour. Let the equations be
\begin{align}
	\frac{\partial u}{\partial t}=\rates(u,p)
	\qquad\mathrm{where}\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}

In the context of the differential equations, a co-dimension one bifurcation can be defined by a set of conditions on the determinant of the Jacobian $\Det$. The determinant of the Jacobian quantifies the rate at which trajectories in a local patch of state-space $u\in\Reals^N$ converge or diverge. The determinant approaching zero means that the dynamics of the system is slowing down, which is an important indicator for the onset of a transition between qualitative behaviours. Furthermore, the slowing down must necessarily be followed by breakdown of stability; for this to be true it is sufficient to require that the determinant cross zero with a finite slope, meaning that its directional derivative along the diagram $\frac{d}{ds}\Det$ is not zero. The set of predicted values for the control condition $\predictions(\theta)\subset\Reals$ at which bifurcations occur are defined as
\begin{align}
	\predictions(\theta):=\left\{\,
	p\,\,\exists\,\,u:\,\,\rates(u,p)=0,\,\,\Det=0,
	\,\, \frac{d}{ds}\Det\neq0
	\,\right\}
	\label{eq:predictions}
\end{align}
 Figure \ref{fig:minimal-models} shows minimal models that explore the space of saddle-nodes $\rates(u,p) = p + \theta_{1}u+\theta_{2}u^3$ and pitchforks $\rates(u,p) = \theta_{1} + p u+\theta_{2}u^3$. Indeed the predictions $\predictions(\theta)$ are defined by zero crossings in the determinant with a finite slope. The location of these crossings in general may not match the targets $\targets$.

For a given set of parameters $\theta$ one could compute the set of predicted bifurcations $\predictions(\theta)$ using parameter continuation methods \cite{Veltz2019PseudoArcLengthContinuation.jl,Farrell2016TheDiagrams}. Our goal is to find optimal parameters $\theta^*$ that match predictions $\predictions(\theta^*)$ to specified targets $\targets$. We must design a suitable cost function $\loss$ so that
\begin{align}
    \theta^* := \mathrm{argmin}_{\theta} \loss(\theta|\targets)
    \label{eq:optimal-theta}
\end{align}
The optimal $\theta^*$ is not expected to always be unique, but is in general a manifold representing the space of qualitatively equivalent models. The cost function $\loss$ must have some distance measure between the two sets $\predictions(\theta)$ and $\targets$ with a bias that encourages equivalent set sizes $|\predictions(\theta)|=|\targets|$. This is especially important in the case where there are no predictions $|\predictions(\theta)|=0$.

\begin{figure}
\centering
\setlength\unitlength{1cm}
{\phantomsubcaption\label{fig:saddle-node}}
{\phantomsubcaption\label{fig:pitchfork}}
\includegraphics[width=6cm]{saddle-node}
\begin{picture}(0,0) \put(-5.8,10){\subref{fig:saddle-node}} \end{picture}
\includegraphics[width=6cm]{pitchfork}
\begin{picture}(0,0) \put(-5.8,10){\subref{fig:pitchfork}} \end{picture}
\caption{Saddle-node model \ref{fig:saddle-node} $\rates(u,p) = p + \theta_{1}u+\theta_{2}u^3$ and pitchfork model \ref{fig:pitchfork} $\rates(u,p) = \theta_{1} + p u+\theta_{2}u^3$ with $\theta=(5/2,-1),(1/2,-1)$ and targets $\targets=\{-1,1\},\{0\}$ respectively. Lighter shades indicate the determinant crossing zero at locations $\predictions(\theta)$ giving rise to unstable solutions}
\label{fig:minimal-models}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=7cm]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars}
    \label{fig:measure}
\end{figure}

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. We expect this term to be some mean over the targets $\targets$ and predictions $\predictions(\theta)$. We choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target in cases where $|\predictions|\geq|\targets|$. This term only zero when each target is matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:error}
\end{align}
We can see from Figures \ref{fig:minimal-models} and definitions \eqref{eq:predictions} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$. Using these quantities we can define a positive semi-definite measure $\measure(u,p)$ of zero crossings in the determinant along the bifurcation curve. Figure \ref{fig:measure} demonstrates this measure is maximal at bifurcation at locations $p(\theta)$ and has finite gradients in non-bifurcating regimes. The total bifurcation measure $\Psi(\theta)$ is defined as
\begin{align}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(u,p)\,\mathrm{d}u\mathrm{d}p
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}u\mathrm{d}p
    }
    \quad\mathrm{where}\quad
    \measure(u,p):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{align}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}u\mathrm{d}p$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$. The measure $\measure(u,p)$ is one at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in the directional derivative $\frac{d}{ds}\Det$ going to zero and hence the measure $\measure(u,p)$ going to zero for each turning point. While the calculation of the determinant is straightforward, its directional derivative is not. Fortunately, a tangent field representation of the bifurcation curve allows us to compute this derivative implicitly; see Appendix \ref{appendix:tangent-fields} for details.

As the determinant $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(u,p)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(u,p)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations. The measure $\measure(u,p)$ is also invariant under certain classes of transformations on $\rates(u,p)$ which do not change bifurcation locations, proofs of which are provided in Appendix \ref{appendix:measure}. The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \frac{1}{|\targets|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:loss}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that in while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. We can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ and total measure $\Psi(\theta)$
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation curve $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:space-curve}.

The gradient of the bifurcation points $\frac{\partial p}{\partial\theta}$ is found by application of the implicit function theorem to a vector function $G_{\theta}:\Reals^{N+1}\rightarrow\Reals^{N+1}$ whos components represent the two constraints $\rates(u,p)=0$ and $\Det=0$. By following a similar strategy to that used by implicit layers \cite{Look2020DifferentiableLayers} we yield an $(N+1)\times M$ Jacobian representing a deformation field \cite{Jos2011OnSurface} for each $\theta$ direction. The gradient we are looking for becomes
\begin{align}
    \frac{\partial p}{\partial\theta} = -\hat{p}\cdot\left.\frac{\partial G_{\theta}}{\partial (u,p)}^{-1}
    \frac{\partial  G_{\theta}}{\partial\theta}\right|_{G_{\theta}(u,p)=0}
    \quad\mathrm{where}\quad
    G_{\theta}(u,p):=\begin{bmatrix}\rates(u,p)\\\Det\end{bmatrix}
    \label{eq:gradient-bifurcation}
\end{align}
Here $\hat{p}$ is a unit vector in $(u,p)\in\Reals^{N+1}$ that picks out the deformations along the $p$-direction. If we wanted to place the bifurcation at target steady state $u'$ as well as target control condition $p'$ we would use the full $(N+1)\times M$ deformation matrix. Calculation of this matrix involves inverting an $(N+1)\times(N+1)$ Jacobian $\frac{\partial G_{\theta}}{\partial(u,p)}$. The determinant of this Jacobian goes to zero in the degenerate case where $\frac{d}{ds}\Det=0$. This further justifies our choice for measure $\Psi(\theta)$ which discourages the degenerate case.

The cost function is piece-wise smooth and differentiable with undefined gradients only in parameter contours where the number of predictions $|\predictions|$ changes; this is when $\Psi(\theta)$ is undefined and the inverse of $\frac{\partial G_{\theta}}{\partial (u,p)}$ does not exist. Given a set of solutions to $\rates(u,p)=0$ and locations $p(\theta)$ the gradient $\frac{\partial\loss}{\partial\theta}$ can evaluated using automatic differentiation methods without needing to back-propagate through the solver that obtained the level set $\rates(u,p)=0$ in the forward pass.
